---
output: github_document
---

This article was written following new functionality added to the `{stplanr}` package by Josiah Parry, in [this pull request (`#540`)`)](https://github.com/ropensci/stplanr/pull/540).
As described issue [`#539`](https://github.com/ropensci/stplanr/issues/539), the original implementation was slower than needed.

This short article describes what the function does, benchmarks alternative implementation, and concludes with some thoughts about the `{stplanr}` package.
It was written straight after the new functionality was implemented, so we'll start by installing that version of the package:

```{r}
remotes::install_github(repo = "ropensci/stplanr", ref = "39858b6")
```

Loading the package allows us to run the function.

```{r}
library(stplanr)
```

`mats2line()` is a function that takes 2 sets of coordinates and converts them into linestrings.
A simple example is shown function's online [documentation](https://docs.ropensci.org/stplanr/reference/mats2line.html) hosted by community science support organsiation rOpenSci:

```{r}
m1 <- matrix(c(1, 2, 1, 2), ncol = 2)
m2 <- matrix(c(9, 9, 9, 1), ncol = 2)
l <- mats2line(m1, m2)
class(l)
l
lsf <- sf::st_sf(l, crs = 4326)
class(lsf)
plot(lsf)
```


The previous version of `mats2line()` as as follows, which shows how it works (see the new implementation in [source code](https://github.com/ropensci/stplanr/blob/39858b69ba1f4ea978a46aec3df0086b9304ce24/R/line_via.R#L19-L47)):

```{r}
mats2line_old <- function(mat1, mat2, crs = NA) {
  l <- lapply(1:nrow(mat1), function(i) {
    mat_combined <- rbind(mat1[i, ], mat2[i, ])
    sf::st_linestring(mat_combined)
  })
  if(is.na(crs)) {
    sf::st_sfc(l)
  } else {
    sf::st_sfc(l, crs = crs)
  }
}
```

The new function is as follows:

```{r}
mats2line
```

We can check the 2 results are similar as follows:

```{r}
l_old <- mats2line_old(m1, m2)
waldo::compare(l_old, l)
```

As shown, the 2 results are identical.

Let's do a quick benchmark:

```{r}
bench::mark(
  mats2line_old(m1, m2),
  mats2line(m1, m2)
)
```

From that you may think there's no benefit to speeding things up.
But, when you look at the benchmark results for larger matrices, you can see the difference (as shown in the initial issue [`#539`](https://github.com/ropensci/stplanr/issues/539):

```{r}
m1 <- matrix(rnorm(20000), ncol = 2)
m2 <- matrix(runif(20000), ncol = 2)
bench::mark(
  mats2line_old(m1, m2),
  mats2line(m1, m2)
)
```



<!-- 

```{r, eval=FALSE}
remotes::install_cran("rsgeo")
```


```{r, eval=FALSE}
packageVersion("rsgeo")
``` 


The new implementation uses the `{rsgeo}` package by default, which can be installed as follows:

If this is the first time installing the package and you have Rust installed, it may take some time.
This is because the package is compiled from source code, which is a bit slower than installing a pre-compiled binary.

You can check which version of the package you have installed as follows (`{rsgeo}` version 0.1.6 or later is required):
-->

