---
title: "Introducing stplanr"
author: "Robin Lovelace"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing stplanr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(stplanr)
library(sp) # needed for geographical objects
library(ggmap) # needed for plotting data
library(rgdal) # required for spTransform
library(RCurl)
```


## Introduction

Common tasks in transportation research and planning include:

- Access and processing of data on transport infrastructure and behaviour
- Analysis and visualisation of origin-destination flow data
- Allocation of flows to the transport (road) network
- Development of models to estimate transport behaviour

**stplanr** facilitates each of these tasks with an integrated 
series of tools and example datasets. The 'sustainable' in the
package name ("sustainable transport planning with R")
refers to both the focus on active travel
and the aim for longevity and accessibility in the software.
Transport planning is notoriously reliant on 'black boxes' and
the same applies to scientific research into transport systems
[@Waddell2002].

**stplanr** seeks to address these issues. After the package
has been installed (see the package's
[README](https://github.com/Robinlovelace/stplanr) for details),
it can be loaded in with `library()`:

```{r, message=FALSE}
library(stplanr)
```

## Accessing and converting data

Transport data is often provided in origin-destination ('OD')
form, either as a matrix or (more commonly) a long table
of OD pairs. An example of this type of raw data is provided
below (see `?flow` to see how this dataset was created).  

```{r}
data("flow", package = "stplanr")
head(flow[c(1:3, 12)])
```

Although the flow data displayed above describes movement over
geographical space, it contains no explicitly geographical
information. Instead, the coordinates of the origins and
destinations are linked to a separate geographical dataset
which also must be loaded to analyse the flows. This is a
common problem solved by the function `gFlow2Lines`.
The geographical data is a set of points representing centroids
of the origin and destinations, saved as a
`SpatialPointsDataFrame`. Geographical data in R is best
represented as such `Spatial*` objects, which use the
`S4` object engine. This explains the close integration of
**stplanr** with R's spatial packages, especially **sp**, which
defines the `S4` spatial object system.

```{r}
data("cents", package = "stplanr")
as.data.frame(cents[1:3,-c(3,4)])
```

We use `gFlow2Lines` to combine `flow` and `cents`, to join
the former to the latter. We will visualise the
`l` object created below in the next section. 

```{r}
l <- gFlow2line(flow = flow, zones = cents)
```

The data is now in a form that is much easier to analyse. We can plot the
data with the command `plot(l)`, which was not possible before. Because the
`SpatialLinesDataFrame` object also contains data per line, it also helps
with visualisation of the flows, as illustrated below.

```{r}
plot(l, lwd = l$All / 10)
```

## Allocating flows to the transport network

A common problem faced by transport researchers is network
allocation: converting the 'as the crow flies' lines illustrated in the
figure above into routes. These are the complex, winding
paths that people and
animals make to avoid obstacles such as buildings and to make the journey
faster and more efficient (e.g. by following the road network).

This is difficult (and was until recently near impossible using free software)
because of the size and complexity of transport networks, the complexity
of realistic routing algorithms and need for context-specificity in the routing
engine. Inexperienced cyclists, for example, would take a very different route
than a heavy goods vehicle.

**stplanr** tackles this issue by using 3rd party APIs to provide
route-allocation. The `gLines2Cyclepath` function allocates straight line
routes to the transport network using the
[CycleStreets.net API](http://www.cyclestreets.net/api/) (you must request
an API key for the function to work):

```{r}
# test the connection
x = try(download.file("http://www.google.com", "test.html"), silent = T) 
offline <- grepl("Error", x)

if(nchar(Sys.getenv('CYCLESTREET')) == 16 & !offline){
  rf <- gLines2CyclePath(l = flowlines)
  rq <- gLines2CyclePath(l = flowlines, plan = "quietest")
} else{
  data("routes_fast")
  data("routes_slow")
  rf <- routes_fast
  rq <- routes_slow
}

plot(l)
lines(rf, col = "red")
```

The above code sends requests to CycleStreets.net and saves the results.
By changing the `plan` argument of `gLines2Cyclepath()`, we can download
routes that are more suitable for people prioritising speed, quietness or
a balance between speed and quietness. The plots of `l` and `rf` illustrate
how the spatial classes that the flow data are converted to are easy to
visualise. Using such data objects, generated by **stplanr**, in combination
with additional packages for creating maps in R
[e.g. see @Cheshire2015; @Kahle2013], will further enhance the utility of
the package's functions. Route-allocated flows are also useful for 
analysis.

## Analysis of origin-destination flow data

Route-allocated lines allow estimation of *route distance* and 
*cirquity* (route distance divided by Euclidean distance).
These variables can help model the rate of flow between origins and
destination, as illustrated in the next section. The code below demonstrates
how objects generated by **stplanr** can be used to estimate route distance.

```{r}
lgb <- spTransform(l, CRSobj = CRS("+init=epsg:27700"))
l$d_euclidean <- rgeos::gLength(lgb, byid = T)
l$d_fastroute <- routes_fast@data$length
plot(l$d_euclidean, l$d_fastroute,
  xlab = "Euclidean distance", ylab = "Route distance")
abline(a = 0, b = 1)
abline(a = 0, b = 1.2, col = "green")
abline(a = 0, b = 1.5, col = "red")
```

The plot illustrates the expected strong correlation between
Euclidean and fastest route distance. However, some OD pairs
have a proportionally higher route distance than others.
This is illustrated
by distance from the black line in the above plot, which represents circuity. 

The difference between
different types of route is shown in the figure below.
'Fast' and 'quiet' routes (red and green, respectively)
are clearly allocated to the transport network
(the package 'leaflet' was used to generate the basemap).


```{r, echo=FALSE, warning=FALSE, message=FALSE}
if(offline){
  print("Figure not shown: requires internet connection")
} else {
  n <- 30
  lf <- fortify(l[n,])
  rff <- fortify(rf[n,])
  rqf <- fortify(rq[n,])
  bb <- bbox(rf[n,])
  bb <- make_bbox(lon = bb[1,], lat = bb[2,], f = 0.5)
  ggmap(ggmap = get_map(location = bb)) +
    geom_path(aes(x = long, y = lat), data = lf) + 
    geom_path(aes(x = long, y = lat), data = rqf, colour = "red") + 
    geom_path(aes(x = long, y = lat), data = rff, colour = "green") +
    theme_nothing()
}
```


## Developing models of travel behaviour

There are many ways of estimating flows between origins and destinations,
including spatial interaction models, the four-stage transport model
and 'distance decay'. **stplanr** aims eventually to facilitate
creation of many types of flow model. 

At present (May 2015) the only functions to assist with this are
the `dd_*` functions, simple commands to create distance decay curves.
Distance decay is an especially important concept for sustainable transport
planning due to physical limitations on the ability of people to walk and
cycle large distances [@Iacono2010].

Let's explore the relationship between distance and the proportion of trips
made by walking, using the same object `l` generated by **stplanr**.

```{r}
l$pwalk <- l$On.foot / l$All
plot(l$d_euclidean, l$pwalk, cex = l$All / 50,
  xlab = "Euclidean distance (m)", ylab = "Proportion of trips by foot")
```

Based on the figure there is a clear negative relationship
between distance of trips and the proportion of those trips made by walking.
This is unsurprising: beyond a certain distance (around 1.5km according
the the data presented in the figure above) walking may a rather slow and
and many people would concider travelling by bike instead!
According to the academic literature, this 'distance decay' is non-linear
and there have been a number of functions proposed to fit to distance decay
curves [@Martinez2013]. From the range of options we test below just two forms.
We will compare the ability of linear and log-square-root functions
to fit the data contained in `l` for walking.

```{r}
lm1 <- lm(pwalk ~ d_euclidean, data = l@data, weights = All)
lm2 <- lm(pwalk ~ d_fastroute, data = l@data, weights = All)
lm3 <- glm(pwalk ~ d_fastroute + I(d_fastroute^0.5), data = l@data, weights = All, family = quasipoisson(link = "log"))
```

The results of these regression models can be seen using `summary()`.
The results for such a small dataset are not particularly interesting and,
surprisingly, Euclidean distance seems to be a better predictor of
walking that route distance. The results are displayed in the graphic below.

```{r, echo=FALSE, eval=FALSE}
summary(lm1)
summary(lm2)
summary(lm3)
```


```{r}
plot(l$d_euclidean, l$pwalk, cex = l$All / 50,
  xlab = "Euclidean distance (m)", ylab = "Proportion of trips by foot")
l2 <- data.frame(d_euclidean = 1:5000, d_fastroute = 1:5000)
lm1p <- predict(lm1, l2)
lm2p <- predict(lm2, l2)
lm3p <- predict(lm3, l2)
lines(l2$d_euclidean, lm1p)
lines(l2$d_euclidean, exp(lm2p), col = "green")
lines(l2$d_euclidean, exp(lm3p), col = "red")
```

The purpose of the `dd*` functions is to provide easy-to-remember commands for
operationalising the formulae used in regression models of distance decay.
Taking the final model, for example, we can reproduce the curve
presented in the figure above using `dd_logsqrt()`:

```{r}
(coefs <- coef(lm3))
p <- dd_logsqrt(1:5000, a = exp(coefs[1]), b1 = coefs[2], b2 = coefs[3])
plot(1:5000, p)
lines(l2$d_euclidean, exp(lm3p), col = "red")
```

## Future plans

The final thing to say about this package is that it is work-in-progress
and we hope to add more functionality, including tools for creating
spatial interaction models, more functions to access routing APIs and
improved functionality for accessing OpenStreetMap data.

If you would like to contribute or request new features, see
https://github.com/Robinlovelace/stplanr

## References

Cheshire, James, and Robin Lovelace. 2015. “Spatial data visualisation with R.” In Geocomputation, edited by Chris Brunsdon and Alex Singleton, 1–14. SAGE Publications. https://github.com/geocomPP/sdv

Iacono, Michael, Kevin J. Krizek, and Ahmed El-Geneidy. 2010. “Measuring non-motorized accessibility: issues, alternatives, and execution.” Journal of Transport Geography 18 (1). Elsevier Ltd: 133–40. doi:10.1016/j.jtrangeo.2009.02.002

Kahle, D, and Hadley Wickham. 2013. “ggmap: Spatial Visualization with ggplot2.” The R Journal 5: 144–61. http://stat405.had.co.nz/ggmap.pdf

Martínez, L. Miguel, and José Manuel Viegas. 2013. “A new approach to modelling distance-decay functions for accessibility assessment in transport studies.” Journal of Transport Geography 26: 87–96. doi:10.1016/j.jtrangeo.2012.08.018

Waddell, Paul. 2002. “UrbanSim: Modeling urban development for land use, transportation, and environmental planning.” Journal of the American Planning Association 68 (3). Taylor & Francis: 297–314

