---
title: "Introducing stplanr"
author: "Robin Lovelace"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing stplanr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Common tasks in transportation research and planning include:

- Access and processing of data on transport infrastructure and behaviour
- Analysis and visualisation of origin-destination flow data
- Allocation of flows to the transport (road) network
- Development of models to estimate transport behaviour

**stplanr** facilitates each of these tasks with an integrated 
series of tools and example datasets. The 'sustainable' in the
package name ("sustainable transport planning with R")
refers to both the focus on active travel
and the aim for longevity and accessibility in the software.
Transport planning is notoriously reliant on 'black boxes' and
the same applies to scientific research into transport systems.

**stplanr** seeks to address these issues. After the package
has been installed (see the package's
[README](https://github.com/Robinlovelace/stplanr) for details),
it can be loaded in with `library()`:

```{r, message=FALSE}
library(stplanr)
library(sp) # needed for geographical objects
```


## Accessing and converting data

Transport data is often provided in origin-destination ('OD')
form, either as a matrix or (more commonly) a long table
of OD pairs. An example of this type of raw data is provided
below (see `?flow` to see how this dataset was created).  

```{r}
data("flow", package = "stplanr")
head(flow[c(1:3, 12)])
```

Although the flow data displayed above describes movement over
geographical space, it contains no explicitly geographical
information. Instead, the coordinates of the origins and
destinations are linked to a separate geographical dataset
which also must be loaded to analyse the flows. This is a
common problem solved by the function `gFlow2Lines`.
The geographical data is a set of points representing centroids
of the origin and destinations, saved as a
`SpatialPointsDataFrame`. Geographical data in R is best
represented as such `Spatial*` objects, which use the
`S4` object engine. This explains the close integration of
**stplanr** with R's spatial packages, especially **sp**, which
defines the `S4` spatial object system.

```{r}
data("cents", package = "stplanr")
as.data.frame(cents[1:3,-c(3,4)])
```

We use `gFlow2Lines` to combine `flow` and `cents`, to join
the former to the latter. We will visualise the
`l` object created below in the next section. 

```{r}
l <- gFlow2line(flow = flow, zones = cents)
```

The data is now in a form that is much easier to analyse. We can plot the
data with the command `plot(l)`, which was not possible before. Because the
`SpatialLinesDataFrame` object also contains data per line, it also helps
with visualisation of the flows, as illustrated below.

```{r}
plot(l, lwd = l$All / 10)
```

## Allocating flows to the transport network

A common problem faced by transport researchers is network
allocation: converting the 'as the crow flies' lines illustrated in the
figure above into routes. These are the complex, winding
paths that people and
animals make to avoid obstacles such as buildings and to make the journey
faster and more efficient (e.g. by following the road network).

This is difficult (and until recently near impossible using free software)
because of the size and complexity of transport networks, the complexity
of realistic routing algorithms and need for context-specificity in the routing
engine. 




```{r}
data("routes_fast")
plot(l)
lines(routes_fast, col = "red")
```











