---
title: "Origin-destination data in R: examples with stplanr"
output: rmarkdown::html_vignette
author: "Robin Lovelace"
vignette: >
  %\VignetteIndexEntry{stplanr vignette 2: origin-destination data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography:
 - references.bib
 - stplanr-citation.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

# Introduction: what is OD data?

Origin-destination (OD) data is data that has not just one geographic location, but two.
As the name suggests, OD data represents movement through geographic space, from an origin (O) to a destination (D).
Despite the rather dull name, OD datasets are a vital part of the modern world, underpinning the transport models that influence current *and future* systems.
If you, like many others, are frustrated by the car dominated nature of most cities, OD datasets should be of great interest: they underlied the car-centric transport models that were helped inform and justify the vast expansion of highways that happened in the 20^th^ Century [@boyce_forecasting_2015].

If you are under any illusion that this healthy outcome of 20^th^ Century transport planning, consider this: roads are now the largest cause of death and injury for young people worldwide, killing and maiming more than more than 1 million and 10 million people each year according to data from the [World Health Organization](https://www.who.int/violence_injury_prevention/road_safety_status/2018/en/).
Even ignoring other problems such as air pollution, an obesity crisis driven by physical inactivity and climate change, it is clear that current transport systems, and the models that have helped to design them, are unsustainable.
There are other reasons why transport data analysis is important, as outlined in the 'stplanr paper' [@lovelace_stplanr_2018].

The purpose of this paper is to provide an introduction to OD paper, with examples based on data and functions from the stplanr package.
OD data can be accessed from a range of sources (we will see code that downloads millions of OD pairs later in this vignette).
Generally you will need to do some 'data carpentry' before the OD data is 'clean' and ready for analysis.
This vignette will not cover cleaning OD, we assume you know R and come with clean (if not 'tidy') data!

In simple terms OD data looks like this:


```{r setup, message=FALSE}
library(stplanr)
library(dplyr)
od = pct::wight_od %>% 
  select(-matches("rail|name|moto|car|tax|home")) %>% 
  top_n(n = 5, wt = all)
```

```{r}
(od_inter = od %>% filter(geo_code1 != geo_code2))
(od_intra = od %>% filter(geo_code1 == geo_code2))
```


# Representations of OD data


## Origin-destination pairs (long form)

Perhaps the most common and straightforward representation of OD data is a 'long thin' data frame.
This is increasingly the format used by official statistical agencies, including the UK's Office for National Statistics (ONS), who provide origin destination data as a `.csv` file.
Typically, the first column is the zone code of origin and the second column is the zone code of the destination, as shown in the code chunk below, which creates an object called `od_data_sample` using the [`pct` package](https://itsleeds.github.io/pct/) (chunk not evaluated for speed):

```{r, eval=FALSE}
remotes::install_github("ITSLeeds/pct")
library(pct)
od_data_all = pct::get_od()
sel_local = 
  od_data_all$geo_code1 %in% cents_sf$geo_code &
  od_data_all$geo_code2 %in% cents_sf$geo_code 
od_data_sample = od_data_all[sel_local, ]
```

```{r, echo=FALSE, eval=FALSE}
usethis::use_data(od_data_sample)
# aim: get top flows by car use multiplied by distance
# subset flows with more than n people driving:
od_cars = od_data_all[od_data_all$car_driver >= 50, ]
cents_ew = pct::get_centroids_ew()
od_cars = od_cars[
  od_cars$geo_code1 %in% cents_ew$msoa11cd &
  od_cars$geo_code2 %in% cents_ew$msoa11cd ,
  ]
desire_lines_cars = od2line(od_cars, cents_ew)
plot(desire_lines_cars[1:999, ])
desire_lines_cars$euclidean_distance_m = as.numeric(sf::st_length(desire_lines_cars)) / 1000
desire_lines_cars$car_km = desire_lines_cars$car_driver * desire_lines_cars$euclidean_distance_m
lines_cars = dplyr::top_n(desire_lines_cars, 20000, car_km)
summary(lines_cars$car_driver)
plot(lines_cars["car_km"])

saveRDS(lines_cars, "lines_cars.Rds")
piggyback::pb_upload("lines_cars.Rds")
```

Data in this form usually has at least one other colum, representing the amount of travel between the zones.
Additional columns represent addition attributes, e.g. a breakdown of flow by time, mode of travel, type of person, or trip purpose.
As shown in the output of printing the sample dataset, the dataset we have downloaded contains column names representing mode of travel (car, bicycle etc):

```{r}
names(od_data_sample)
```

Each 'mode' column contains an integer, while additional columns contain characters, as demonstrated by printing the data frame:

```{r}
od_data_sample
```


## Origin destination matrices

## Desire lines

The previous two representations of OD data are 'implicitly geographic': their coordinates are not contained in the data, but associated with another data object, typically a zone or a zone centroid.
This is clearly problematic because it makes OD data less modular and self standing.
A useful way of representing OD data that overcomes this issue is `desire_line` data: geographic lines between origin and destination, with information on flow levels between the two.

This is represented in the file `lines_cars.Rds`, representing the top 20,000 desire lines at the MSOA-MSOA level in England and Wales by the number of car km used for travel to work, which can be downloaded, read-in and plotted as follows:


```{r, out.width="100%", warning=FALSE, eval=FALSE}
u = "https://github.com/ropensci/stplanr/releases/download/0.2.9/lines_cars.Rds"
f = file.path(tempdir(), "lines_cars.Rds")
download.file(u, f)
lines_cars = readRDS(f)
plot(lines_cars["car_km"], lwd = lines_cars$car_km / 1000)
```

Based on the estimate of the average energy use per km being 2.5 MJ, and that these return trips are made on average 200 times per year, with a circuity of 1.3, we can estimate the total energy use of the 'high energy commutes' as follows:

```{r, eval=FALSE}
sum(lines_cars$car_km * 2.5 * 200) / 1e9
```

That represents ~10 petajoules (PJ), only for the top 20,000 most energy intensive commutes.
That may seem like alot, but represents only a fraction of the UK's total energy use of [~200 Mtoe](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/729451/DUKES_PN.pdf) (8400 PJ).


```{r, echo=FALSE, eval=FALSE}
# out-takes and test code
u = "http://www.rita.dot.gov/bts/sites/rita.dot.gov.bts/files/publications/commodity_flow_survey/2007/zip/origin_destination_files.zip"
f =  file.path(tempdir(), "origin_destination_files.zip")
download.file(u, f)
unzip(f)
```

